---
title: "owner-analysis"
author: "Andrew L"
format: html
editor: visual
toc: true
---

## Purpose

Analyze lobster abundance and size data sets from the Santa Barbara Coastal LTER. This data helps informs scientists about the abundance of California spiny lobster (*Panulirus interruptus*). This data can help make predictions on overall health and quantity of lobster along the Santa Barbara Channel

Reed, Daniel C. *SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012*. Santa Barbara Coastal LTER. https://doi.org/10.6073/pasta/25aa371650a671bafad64dd25a39ee18.

Here we are reviewing the counts (abundance) data set. This data set appears to monitor the lobster within a specific area.

From the Authors:

This dataset contains two tables. 1) Abundance and size data collected annually by divers in late summer before the start of the fishing season at five SBC LTER long term kelp forest study sites: two within MPAs (Naples and Isla Vista) and three outside (Arroyo Quemado, Mohawk and Carpinteria). 2) Fishing pressure, as determined by counting the number of commercial trap floats. Data are collected every two to four weeks during the lobster fishing season (October to March) at nine sites along the mainland, eight of which are also SBC LTER long-term kelp forest study reefs.

See Methods for more information. 

## Data clean up & quality control

First import our R libraries

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(here)
```

We then import the data. For this analysis I will be using the lobster abundance csv file.

```{r}
#load the aata
lobster_abundance <- read_csv(here::here("data/Lobster_Abundance_All_Years_20220829.csv"))
#view the data
head(lobster_abundance, n=5)
summary(head(lobster_abundance,n=10))
list(lobster_abundance)
```

### Exercise 1 Question 1

When the dataset was generated, a NA input for a lobster's size was interpreted as -99999. First we need to correct this value.

```{r}
lobster_abundance <- lobster_abundance %>% 
    mutate(SIZE_MM = na_if(SIZE_MM, -99999))
```

### Exercise 2 Question 2

Next we are excluding lobsters from the NAPL region.

```{r}
not_napl <- lobster_abundance %>% 
    filter(SITE != "NAPL")
```

### Exercise 3 Question 3

We are going to create and store a subset with lobsters at Arroyo Quemado (`AQUE`) AND with a carapace length greater than 70 mm.

```{r}
aque_70mm <- lobster_abundance %>% 
    filter(SITE == "AQUE" & SIZE_MM >= 70)
```

We can group the data based on other factors we may care about. Here we can view the lobster size anchored to it's location.

```{r}
#`group_by() %>% summarize()` practice

max_lobster <- lobster_abundance %>% 
  group_by(SITE, MONTH) %>% 
  summarize(MAX_LENGTH = max(SIZE_MM, na.rm = TRUE))
```

## Exercise: Data Visualization

First we will
