---
title: "Lobster Report"
author: "Andrew L & Maria Mendoza"
format: html
editor: visual
toc: true
---

## Purpose

Analyze lobster abundance and size data sets from the Santa Barbara Coastal LTER. This data helps informs scientists about the abundance of California spiny lobster (*Panulirus interruptus*). This data can help make predictions on overall health and quantity of lobster along the Santa Barbara Channel

Reed, Daniel C. *SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012*. Santa Barbara Coastal LTER. https://doi.org/10.6073/pasta/25aa371650a671bafad64dd25a39ee18.

Here we are reviewing the counts (abundance) data set. This data set appears to monitor the lobster within a specific area.

From the Authors:

This dataset contains two tables. 1) Abundance and size data collected annually by divers in late summer before the start of the fishing season at five SBC LTER long term kelp forest study sites: two within MPAs (Naples and Isla Vista) and three outside (Arroyo Quemado, Mohawk and Carpinteria). 2) Fishing pressure, as determined by counting the number of commercial trap floats. Data are collected every two to four weeks during the lobster fishing season (October to March) at nine sites along the mainland, eight of which are also SBC LTER long-term kelp forest study reefs.

See Methods for more information. 

## Data clean up & quality control

First import our R libraries

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(here)
```

We then import the data. For this analysis I will be using the lobster abundance csv file.

```{r}
#load the aata
lobster_abundance <- read_csv(here::here("data/Lobster_Abundance_All_Years_20220829.csv"))

```

## Data analysis for Abundance

I would like to visualize the percentage of all caught lobster based on size:

```{r}
#| label: fig - Total Large vs Small Lobster Caught
#First identify our lobster classifications
lobster_size_lrg <- lobster_abundance %>% filter(YEAR %in% c(2019, 2020, 2021)) %>% mutate(SIZE_BIN = if_else(SIZE_MM <= 70, true = "small", false = "large")) %>%group_by(SITE, SIZE_BIN) %>%  summarize(COUNT = n()) %>% drop_na()

#visualize
ggplot(data = lobster_size_lrg, aes(x = SITE, y = COUNT, fill = SIZE_BIN)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("large" = "#E76F51","small" = "#2A9D8F")) +
  labs(x = "Site",y = "Percent of total lobsters",fill = "Size bin")
ggsave(here::here("figs/total_lobsterpercent.png"))
```

I would also like to view this same data by year. To do this, i would first need to refilter my object by year.\

```{r}
#Filter by size and year
lobster_size_lrg_year <- lobster_abundance %>%
  filter(YEAR %in% 2012:2022) %>%
  mutate(SIZE_BIN = if_else(SIZE_MM <= 70, "small", "large"),YEAR = factor(YEAR)) %>%group_by(YEAR, SITE, SIZE_BIN) %>%summarize(COUNT = n(), .groups = "drop") %>%drop_na()

#plot it
ggplot(lobster_size_lrg_year,
       aes(x = YEAR, y = COUNT, fill = SIZE_BIN)) +
  geom_col(position = "fill") +
  facet_wrap(~ SITE) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Year",y = "Percent of lobsters",fill = "Size bin") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),strip.text = element_text(size = 12),legend.position = "right")
ggsave(here::here("figs/total_lobsterbyyear.png"))
```

lobster_abundance %\>% filter(YEAR %in% c(2019, 2020, 2021)) %\>% count(MONTH)
